Pillar 1 - Sequential Pattern Mining

### 2019-01-29 Pillar 1 - Sequential Pattern Mining.Rmd
Integrates not taken courses into d_transactions

    item_PF = paste(`Course ID`, PF, sep = "_"),
    item_HL = paste(`Course ID`, HL, sep = "_")
    ) %>%
  rename(
    item = `Course ID`
  )
```

@@

#
# Making transactions
transactions    <- make_transaction(item = item   )
transactions_PF <- make_transaction(item = item_PF)
transactions_HL <- make_transaction(item = item_HL)

#
# Checking transactions
inspect(head(transactions, 10))

@@

#
# Checking sequences
inspect(head(sequences, 10))

@@

  AR %>%
    as("data.frame") %>%
    filter(
      count >= 1
    ) %>%
    mutate(
      rules = str_remove(rules, pattern = "_fail"),
      rules = str_remove(rules, pattern = "_low"),
      lhs = substr(rules, 2, 8),
      rhs = substr(rules, 15, 21),
      `rhs.support` = confidence / lift
      ) %>%
    mutate_at(
      c("support", "lhs.support", "confidence", "rhs.support", "lift"),
      funs(round(., 5))
      ) %>%
    select(
      lhs, rhs,
      support, count, lhs.support,
      confidence, rhs.support,
      lift
      ) %>%
    arrange(
      desc(count)
    )

@@

# creating vector of fail course
course_id_low <- d_transactions %>%
  filter(HL == "low") %>%
  distinct(`item_HL`)

@@

#
# Save association rules
save(AR, AR_PF, AR_HL,
     file = "Output/AS.RDATA")

@@

rm(clean_AR, my_apriori,
   transactions, transactions_PF, transactions_HL,
   course_id_fail, course_id_low
   AR, AR_PF, AR_HL)

@@

#TRASH:
##Condensing grades

```{r}
tmp_distributed_transcript <- d_transcript %>%
  filter(!is.na(Grade), Grade>=0) %>%
  mutate(Round_grade = paste("g", round(Grade,0),sep=""),
         Values = T)
#This is the transcript with cumulative error gap.
distributed_transcript <- spread(tmp_distributed_transcript, key = "Round_grade", value = Values #, fill = F
               )%>%
  mutate(g1  = g0|g1,
         g2  = g1|g2,
         g3  = g2|g3,
         g4  = g3|g4,
         g5  = g4|g5,
         g6  = g5|g6,
         g7  = g6|g7,
         g8  = g7|g8,
         g9  = g8|g9,
         g10 = g9|g10,
         ) %>%
  gather(key="Gr_string", value="Gr_bool", g0, g1, g2, g3, g4,g5,g6,g7,g8,g9,g10, na.rm=T) %>%
  select(-Gr_bool )%>%
  separate(Gr_string, into=c("Trash", "Grade_new"), sep="g")%>%
  select(-Trash) %>%
  mutate(Grade= as.numeric(Grade_new))%>%
  select(-Grade_new)%>%
  filter(!str_detect(Period, " to "))%>% #Crutch FILTER
  mutate(Item=paste(`Course ID`,Grade, sep="_"))%>%
  mutate( Time = paste(Year_numerical, Period, sep = ""),
          Time = as.double(Time))%>%
  select(`Student ID`, Item, Time) %>%
  arrange(`Student ID`, Time) %>% #Note that integer identifiers must be positive and that transactions must be ordered by sequenceID and eventID.
  group_by(`Student ID`, Time) %>%
  summarize(list_course = list(Item)) %>%
  mutate(TransactionID = paste(`Student ID`, Time, sep = "_"))

@@

Excluding manatory courses.
```{r}

tmp_distributed_transcript <- d_course %>%
  select(`Course ID`, Type, Letters)   %>%
  # Exclude 
  filter(
    Type != "Mandatory",                  # (i) mandatory courses e.g. COR, CAP, etc
    ! Letters %in% c("SKI", "PRO",        # (ii) skills & projects (taken by majority of students)
                     "SAS", "SAH", "SAC") # (iii) courses of semester abroad (uninformative)
    ) %>%

  # Join with transcripts.
  inner_join(d_transcript, by = "Course ID") %>%
  filter(!is.na(Grade), Grade>=0)            %>%
  mutate(Round_grade = paste("g", round(Grade,0),sep=""),
         Values = T)
#This is the transcript with cumulative error gap.

distributed_transcript <- spread(tmp_distributed_transcript, key = "Round_grade", value = Values #, fill = F
               ) %>%
  mutate(g1  = g0|g1,
         g2  = g1|g2,
         g3  = g2|g3,
         g4  = g3|g4,
         g5  = g4|g5,
         g6  = g5|g6,
         g7  = g6|g7,
         g8  = g7|g8,
         g9  = g8|g9,
         g10 = g9|g10,
         ) %>%
  gather(key="Gr_string", value="Gr_bool", g0, g1, g2, g3, g4,g5,g6,g7,g8,g9,g10, na.rm=T) %>%
  select(-Gr_bool )%>%
  separate(Gr_string, into=c("Trash", "Grade_new"), sep="g")%>%
  select(-Trash) %>%
  mutate(Grade= as.numeric(Grade_new))%>%
  select(-Grade_new)%>%
  filter(!str_detect(Period, " to "))%>% #Crutch FILTER
  mutate(Item=paste(`Course ID`,Grade, sep="_"))%>%
  mutate( Time = paste(Year_numerical, Period, sep = ""),
          Time = as.double(Time))%>%
  select(`Student ID`, Item, Time) %>%
  arrange(`Student ID`, Time) %>% #Note that integer identifiers must be positive and that transactions must be ordered by sequenceID and eventID.
  group_by(`Student ID`, Time) %>%
  summarize(list_course = list(Item)) %>%
  mutate(TransactionID = paste(`Student ID`, Time, sep = "_"))

@@

#puting everything into quasi-function
--------------------------------------------------
##to transaction
```{r}
transaction_list_course <- distributed_transcript$list_course
names(transaction_list_course) <- distributed_transcript$TransactionID

transactions <- as(transaction_list_course, "transactions")
transactionInfo(transactions)$sequenceID <- distributed_transcript$`Student ID`
transactionInfo(transactions)$eventID <- distributed_transcript$Time
transactionInfo(transactions)$transactionID <- NULL

#checking results
as(transactions, "data.frame")
```

##Create filter condition.
```{r}
# string_filter <- case_when(quo_name(category)=="Outcome"        ~ "Fail",
#                            quo_name(category)=="Low_grade"      ~ "low score",
#                            quo_name(category)=="Discrete_grade" ~ "5",
#                            quo_name(category)=="None"           ~ "_" #useless filter
#                            )
```
##CSPADE Algorithm:

```{r}
frequent_trajectories <- cspade(transactions, parameter = list(support = 0.1,
                                                               maxsize = 1,
                                                               maxlen = 2 #,
                                                               #mingap=, #gap between consecutive elements
                                                               #maxgap=, #hap between consecutive elements
                                                               #maxwin=  #gap between ANY two elements
                                                                 ), control   = list(verbose = TRUE))

#Inspect results CSPADE
summary(frequent_trajectories)
as(frequent_trajectories, "data.frame")

```
##rule induction:
```{r}
rules <- ruleInduction(frequent_trajectories,
                       confidence = 0, #we need to play with the parameters.
                       control    = list(verbose = TRUE))

  #filter(str_detect(RHS, string_filter)) %>% #FILTER (YES) --> uncomment, (NO)-->comment

support(frequent_trajectories,transactions, control=list(verbose=T))

```
here I'l playing around with the correct confidence:

#correcting lookup for this case:
First we find out raw proportions per course (how many were low grade/high grade, out of a particular course, and the percentage this represents from the total)
```{r}
lookup_table <- d_course %>%

  select(Type, `Course ID`, Letters) %>%

  filter(Type == "Elective",
         ! Letters %in% c("SKI", "PRO"),
         ! Letters %in% c("SAS", "SAH", "SAC")) %>%

  select(`Course ID`) %>%

  left_join(d_transcript, by = "Course ID") %>%
  mutate(
    Outcome = 
           case_when( Grade == -1   ~"Fail", #if we don't do this NG satisfies the condition for pass v
                      Grade  > 5.5 ~"Pass",
                               T~"Fail"),
    Low_grade =
           case_when(Grade == -1 ~"low score", #if we don't do this NG satisfies the condition for pass v
                      Grade > fail_threshold ~"high score",
                               T~"low score"),
    Discrete_grade =
           case_when(Grade <= 5.4             ~ 5.0,
                     between(Grade, 5.5, 6.4) ~ 6.0,
                     between(Grade, 6.5, 7.4) ~ 7.0,
                     between(Grade, 7.5, 8.4) ~ 8.0,
                     between(Grade, 8.5, 9.4) ~ 9.0,
                     Grade >= 9.5             ~ 10.0,
                     T ~ -99999    #this is a control, we should not have any grades here
                     ),
    None= "" # this might not be optimal
         ) %>%

  #mutate(Item = paste(`Course ID`, !!category, sep = "_")) %>% #Change for Outcome/Low_grade/Discrete_grade

  filter(!str_detect(Period, " to ")) %>% #filter courses that spann more than one period

  mutate( Time = paste(Year_numerical, Period, sep = ""),
          Time = as.double(Time)) %>%

  select(`Student ID`, #Item,  #************************************ <- 
         `Course ID`, !!category, Time) %>% # Course ID and Time may not be necessary

  arrange(`Student ID`, Time) %>% #Note that integer identifiers must be positive and that transactions must be ordered by sequenceID and eventID.

  group_by(`Course ID`, !!category) %>% ##this makes no sense we should just group id and low_grade

  summarize(Count = n()) %>%
  #separate(Item, into= c("Course ID", quo_name(category)), sep="_", remove = F) %>%
  ungroup() %>%
  group_by(`Course ID`) %>%
  mutate(Total= sum(Count))%>%
  filter(Low_grade=="low score")%>%
  mutate(Prop_low= Count/Total)
```

```{r}
category
n_students <- length(unique(d_sequence_transactions$`Student ID`))

sequence_rules <- pass_rules_df %>%
  separate(LHS, into=c("LHS Course ID", "LHS Score"), sep="_", remove= FALSE) %>%
  separate(RHS, into=c("RHS Course ID", "RHS Score"), sep="_", remove= FALSE) %>%
  mutate(`RHS Course ID`= substr(`RHS Course ID`, start=4, stop=10))%>% #add more
  left_join(lookup_table, by = c("RHS Course ID"="Course ID")) %>%

  # Compute statistics
  group_by(LHS,`RHS Course ID`) %>%
  mutate(
    Prob_both = sum(support))%>%
  ungroup()%>%
  mutate( 
    Conf_corr = support/Prob_both,
    n_rule    = support * n_students,
    n_both    = Prob_both * n_students,
    lift_diff = Conf_corr - Prop_low,
    lift_corr = Conf_corr / Prop_low
    )

##Puting students into take and not take. 
```{r}
#This df has 731,676 rows:
test <- expand.grid(`Student ID`= unique(d_transcript$`Student ID`),`Course ID`= unique(d_course$`Course ID`)) %>%
  left_join(d_transcript, by=c("Course ID", "Student ID")) %>%
  select(`Course ID`, `Student ID`, Grade, Year_numerical, Period) %>%
  mutate(
    take_course = case_when(
      is.na(Grade)  ~ FALSE,
      T             ~ TRUE
      ),
     PFN = case_when(
       is.na(Grade)  ~ "not",
       Grade >= 5.5  ~ "pass",
       Grade < 5.5   ~ "fail"
       ),
    Item = paste(`Course ID`, take_course, sep="_")
    )

table(test$PFN)

```{r}
test %>%
 mutate(
    Period  = substr(Period, 1, 1),
    eventID = as.numeric(paste(Year_numerical, Period, sep = "")),
    Item = paste(`Course ID`,Grade, sep="_")
    ) %>%
  select(`Student ID`, Item, eventID) %>%
  arrange(`Student ID`, eventID) %>% #Note that integer identifiers must be positive and that transactions must be ordered by sequenceID and eventID.
  group_by(`Student ID`, Time) %>%
  summarize(list_course = list(Item)) %>%
  mutate(TransactionID = paste(`Student ID`, Time, sep = "_"))
```

>>>>

    item_PF = paste(`Course ID`, PF, sep = "_"),
    item_HL = paste(`Course ID`, HL, sep = "_")
    ) %>%
  rename(
    item = `Course ID`
  )
```
### SOFIA---Take not take courses
```{r}
#item and sequenceID

d_transactions_tmp <- expand.grid(sequenceID = unique(d_transactions$sequenceID),item = unique(d_transactions$item)) %>%
  left_join(d_transactions, by = c("sequenceID","item")) %>%
  mutate(
    Taken= case_when(
     is.na(Grade)  ~ "not taken",
      T             ~ "taken"),
    item_Taken= paste(item,Taken, sep = "_")
     )

#ARRANGE PROPERLY

d_transactions <- d_transactions_tmp

rm(d_transactions_tmp)
```

@@

#
# Making transactions
transactions       <- make_transaction(item = item   )
transactions_PF    <- make_transaction(item = item_PF)
transactions_HL    <- make_transaction(item = item_HL)
transactions_Taken <- make_transaction(item=item_Taken)

#
# Checking transactions
#inspect(head(transactions, 10))

@@

#
# Checking sequences
#inspect(head(sequences, 10))

rm(make_sequence)
```

@@

  AR %>%
    as("data.frame") %>%
    filter(
      count >= 1
    ) %>%
    mutate(
      rules = str_remove(rules, pattern = "[{]"),
      rules = str_remove(rules, pattern = "[}]"),
      rules = str_remove(rules, pattern = "[{]"),
      rules = str_remove(rules, pattern = "[}]"), #this is needed 4 times, because it only removes the first instance
      `rhs.support` = confidence / lift
    ) %>%
    separate(rules, into = c("lhs", "rhs"), sep="=>") %>%
    mutate_at(
      c("support", "lhs.support", "confidence", "rhs.support", "lift"),
      funs(round(., 5))
    ) %>%
    select(
      lhs, rhs,
      support, count, lhs.support,
      confidence, rhs.support,
      lift
    ) %>%
    arrange(
      desc(count)
    )

@@

# creating vector of low score course
course_id_low <- d_transactions %>%
  filter(HL == "low") %>%
  distinct(`item_HL`)

@@

AR_Taken <- my_apriori(transactions_Taken)

#
# Save association rules
save(AR, AR_PF, AR_HL, AR_Taken,
     file = "Output/AS.RDATA")

@@

rm(clean_AR, my_apriori,
   transactions, transactions_PF, transactions_HL,
   course_id_fail, course_id_low,
   AR, AR_PF, AR_HL, AR_Taken)

@@

##Not ready to move up:
##Puting students into take and not take. 
```{r}
#This df has 731,676 rows:
test <- expand.grid(`Student ID`= unique(d_transcript$`Student ID`),`Course ID`= unique(d_course$`Course ID`)) %>%
  left_join(d_transcript, by=c("Course ID", "Student ID")) %>%
  filter(
     Type != "Mandatory",                  # (i) mandatory courses e.g. COR, CAP, etc
    ! Letters %in% c("SKI", "PRO",        # (ii) skills & projects (taken by majority of students)
                     "SAS", "SAH", "SAC") # (iii) courses of semester abroad (uninformative)
  )%>%
  select(`Course ID`, `Student ID`, Grade, Year_numerical, Period) %>%
  mutate(
    take_course = case_when(
      is.na(Grade)  ~ "not taken",
      T             ~ "taken"
      ),
     PFN = case_when(
       is.na(Grade)  ~ "not",
       Grade >= 5.5  ~ "pass",
       Grade < 5.5   ~ "fail"
       ),
    Item = paste(`Course ID`, take_course, sep="_")
    )

@@

```{r}
trans <- test %>%
 mutate(
    Period  = substr(Period, 1, 1),
    eventID = as.numeric(paste(Year_numerical, Period, sep = ""))
    ) %>%
   rename(sequenceID = `Student ID`)%>%
  select(sequenceID, eventID, Item) %>%
  make_transaction(item=Item)

@@

```{r}
# Run the apriori algorithm with desired parameters
my_apriori <- function(data, appearance = NULL){

  data %>%
    apriori(
      parameter = list(
        supp = 0,    # min support
        smax = 1,    # max support
        conf = 0,    # min confidence
        minlen = 2,  # min length of 
        maxlen = 2,  # max length of rule
        ext = TRUE,
        arem = "diff",
        aval = TRUE,
        minval = 0
        ),
      appearance = appearance,
      control = list(
        verbose = FALSE
        )
      ) #%>%
    #clean_AR

#
# Generating asocciation rules
AR <- my_apriori(trans)

AR %>%
    as("data.frame") %>%
    filter(
      count >= 1
    )

```
```
